Interisland movements of woodland caribou in Newfoundland

### 2020-02-21 paper/manuscript.Rmd
Update manuscript.Rmd

## Individual variation 
In total, 3 of 29 collared female caribou swam between islands in the Fogo
archipelago. These swimming events were relatively frequent, on average occuring
once every 24 <!--double check--> days during ice-free seasons. However, despite relatively frequent swimming events for these three individuals, we were unable to
identify any swimming events for the other 26 collared females. It remains
unclear why ocean swimming is relatively common among ~10% of our marked
population, while the other 90% never swam. A potential explanation is the Our initial direct observation of
swimming was an adult male, but as we only have GPS data for females, we are
unable to make conclusions on the prevalence of this behaviour among male
caribou. It is theoretically possible that these other females moved rapidly
back and forth between islands within the 2 hour fix rate of our GPS collars,
which we would be unable to detect. We consider this highly unlikely given the
range of residency times for caribou on any given island (min - max: XX - XX).
<!-- What is the typical residency time? Is this a true statement or am I just making this up? -->

The two hypotheses presented above are generally thought to apply uniformly
across populations, but individual caribou could evaluate the costs and benefits
of swimming to another island differently. Potentially those three females,
compared to the other 26, considered swimming less costly or perceived
competition or predation to be of greater concern. The individual attributes
that might influence how animals vary in their assessments of these trade-offs
are unknown, and potential predictors such as age or personality traits are
unknown for these caribou.

## Summary and conclusion
For several decades after their introduction to Fogo Island, the caribou
population did not exceed ~100 individuals (Bergerud & Mercer 1989; Newfoundland
and Labrador Wildlife Division, unpublished data). During the 1990s, population
density reached 300 animals and anecdotal evidence suggests that [only then???]
caribou began to occupy other islands in the archipelago. We surmise that as
competition among conspecifics increased along with population size,
density-dependent habitat selection resulted in expansion of the population to
nearby islands. Following Ideal Free Distribution theory, caribou should swim to
new islands when the average fitness of individuals on the starting island
exceeds the density-fitness equilibrium (Morris 1987). While this is an ultimate
explanation to a series of proximate observations, it is possible for
density-dependent habitat selection to operate at such a fine-scale (Webber &
Vander Wal 2018). Evidence from the Sable Island horse (Equus ferus caballus)
system corroborates this hypothesis (van Beest et al. 2014). For horses,
selection for high quality habitat was strongest at low population densities,
but individuals tended to settle in lower quality habitats as population density
increased, following Ideal Free Distribution theory (van Beest et al. 2014).
Predator avoidance can also contribute to caribou swimming behaviour, but in
this situation it seems a less likely explanation given predation is primarily a
concern for young calves, which are the least likely to engage in this costly
behaviour.
<!-- this last bit is sort of me bullshitting, not sure it's actually defensible?-->

>>>>

## Individual variation 
In total, 3 of 29 collared female caribou swam between islands in the Fogo
archipelago. These swimming events were relatively frequent, on average occuring
once every 24 <!--double check--> days during ice-free seasons. However, despite
relatively frequent swimming events for these three individuals, we were unable
to identify any swimming events for the other 26 collared females. It remains
unclear why ocean swimming is relatively common among ~10% of our marked
population, while the other 90% never swam. Of the three collared individuals,
two were collared on Western Indian Island and the third was collared on Fogo
Island. It is possible that individuals born on, or that live on, Eastern and
Western Indian Island swim are more likely to swim between islands because
inter-island distances are shorter than between Western Indian and Fogo Islands
(Figure 2). Furthermore, it is unclear whether swimming is more common among
males or females. Our initial observation  was of an adult male, but our GPS
data only includes females. It is theoretically possible that these other
females moved rapidly back and forth between islands within the 2 hour fix rate
of our GPS collars, which we would be unable to detect. We consider this highly
unlikely given the range of residency times for caribou on any given island (min
- max: XX - XX). <!--double check-->
<!-- What is the typical residency time? Is this a true statement or am I just making this up? QW: this is something we will look into, I think. -->

The forage limitation and predator avoidance hypotheses are generally thought to
apply uniformly across populations, but individual caribou could evaluate the
costs and benefits of swimming to another island differently. Potentially those
three females, compared to the other 26, considered swimming less costly or
perceived competition or predation to be of greater concern. The individual
attributes that might influence how animals vary in their assessments of these
trade-offs are unknown, and potential predictors such as age or personality
traits are unknown for these caribou.

## Summary and conclusion
For several decades after their introduction to Fogo Island, the caribou
population did not exceed ~100 individuals (Bergerud & Mercer 1989; Newfoundland
and Labrador Wildlife Division, unpublished data). During the 1990s, population
density reached 300 animals and anecdotal evidence suggests that [only then???]
caribou began to occupy other islands in the archipelago. We surmise that as
competition among conspecifics increased along with population size,
density-dependent habitat selection resulted in expansion of the population to
nearby islands. Following Ideal Free Distribution theory, caribou should swim to
new islands when the average fitness of individuals on the starting island
exceeds the density-fitness equilibrium [@Morris_1987]. While this is an ultimate
explanation to a series of proximate observations, it is possible for
density-dependent habitat selection to operate at such a fine-scale (Webber &
Vander Wal 2018). Evidence from the Sable Island horse (Equus ferus caballus)
system corroborates this hypothesis (van Beest et al. 2014). For horses,
selection for high quality habitat was strongest at low population densities,
but individuals tended to settle in lower quality habitats as population density
increased, following Ideal Free Distribution theory (van Beest et al. 2014).
Predator avoidance can also contribute to caribou swimming behaviour, but in
this situation it seems a less likely explanation given predation is primarily a
concern for young calves, which are the least likely to engage in this costly
behaviour.
<!-- this last bit is sort of me bullshitting, not sure it's actually defensible?-->
